Deriving Multi-Headed Projective Dependency Parses \\ from Link Grammar Parses

### 2014-09-21 doc/LinksToDAG.tex
edits from reviewer #1 of EMNLP short submission, including additions to bib

\begin{abstract}
Under multi-headed dependency grammar, a parse is a directed acyclic graph rather than a tree.  Such formalisms can be more syntactically and semantically expressive.  However, 
it is hard to train, test, or improve multi-headed parsers because few multi-headed corpora exist, particularly for the projective case.
% OLD
% investigate the benefit of such parsers or to work on making them faster or more accurate.  
To help fill this gap, we observe that link grammar already produces {\em undirected} projective graphs.  
% OLD
% ... produces parses that are similar to multi-headed projective dependency parses, except that the links are undirected. 
We use Integer Linear Programming to assign consistent directions to the labeled links in a corpus of several thousand parses produced by the Link Grammar Parser, which has broad-coverage hand-written grammars of English as well as Russian and other languages.  We find that such directions can indeed be consistently assigned in a way that yields valid multi-headed dependency parses. The resulting parses in English appear reasonably linguistically plausible, though differing in style from CoNLL-style parses of the same sentences; we discuss the differences.  
\end{abstract}

@@

As recently discussed by \newcite{gomezrodriguez-nivre-2013}, one might impose various requirements on the parse graph:
\begin{itemize}[noitemsep]
\item {\sc Single-Head}: each word has $\leq 1$ parent
\item {\sc Acyclic}: there are no directed cycles
\item {\sc Connected}: each pair of words has a undirected path between them
\item {\sc Reachable}: each word can be reached from 0 by a directed path ($\Rightarrow$ {\sc Connected})
\item {\sc Planar}: edges may not ``cross''\footnote{That is, if there are edges between $i,j$ and between $u,v$, where $i < u < j$, then {\sc Planar} requires $i \leq v \leq j$.}
\end{itemize}
It is common to impose all of these requirements at once, leading to a {\em projective dependency parser} that produces projective trees rooted at 0.\footnote{{\sc Projective} basically means {\sc Planar} + {\sc Reachable}.  Note: It does not prevent 0 from having multiple children.}  However, parsing algorithms can be devised that relax any or all of the requirements \cite{gomezrodriguez-nivre-2013}.

@@

A few past NLP papers have explored multi-headed dependency parsing \cite{buchkromann-2006,mcdonald-pereira-2006,sagae-tsujii-2008,gomezrodriguez-nivre-2013}.  Unfortunately, there seem to be no annotated corpora in this form other than the Danish Dependency Treebank \cite{kromann-2003}; researchers have sometimes converted corpora from other formats such as HPSG.

@@

It seems at first that no one has worked out annotation conventions for {\em projective} multi-headed dependency parsing.  However, this is not quite true.  Link Grammar \cite{SleatorTemperly91} is a grammar-based formalism for projective dependency parsing with {\em undirected} links.  It produces undirected connected planar graphs.  Annotation conventions are implicit in the detailed lexicon for the Link Grammar Parser,\noteJE{check caps}%
\footnote{\url{http://www.abisource.com/projects/link-grammar/dict/introduction.html}.  The 122 link types in the English lexicon are documented at \url{http://www.abisource.com/projects/link-grammar/dict/summarize-links.html}.} 
which specifies for every word a constraint on the {\em sequence} of labeled leftward and rightward edges attached to it.  As remarked by \newcite{eisner-2000-iwptbook}, this is analogous to dependency grammar's use of head automata to constrain a word's sequence of left and right children.  For example, in {\em ``The boy that Jill skipped with fell down,''} the word {\em with} uses a lexical entry that requires it to link to a governing verb to the left, an extracted object farther to the left, and nothing to the right.  Each entry has a hand-assigned cost in \{0,1,2\} and the parser finds the parse of minimum total cost.
% Information about cost is in this message from Linas Vepstas on 2014-02-01: 
%   https://groups.google.com/d/msg/link-grammar/eeJw1Ofgc9U/diqPYSwuFfoJ

@@

However, working directly with the link grammar lexicon format is somewhat tricky.  We also feared that there would not be a perfect solution---for example, because of errors in the lexicon, or linguistically unnatural word sequences not anticipated by the grammar designers.  In this case it would not be clear how to relax our constraints.

>>>>

\begin{abstract}
Under multi-headed dependency grammar, a parse is a connected DAG rather than a tree.  Such formalisms can be more syntactically and semantically expressive.  However, 
it is hard to train, test, or improve multi-headed parsers because few multi-headed corpora exist, particularly for the projective case.
% OLD
% investigate the benefit of such parsers or to work on making them faster or more accurate.  
To help fill this gap, we observe that link grammar already produces {\em undirected} projective graphs.  
% OLD
% ... produces parses that are similar to multi-headed projective dependency parses, except that the links are undirected. 
We use Integer Linear Programming to assign consistent directions to the labeled links in a corpus of several thousand parses produced by the Link Grammar Parser, which has broad-coverage hand-written grammars of English as well as Russian and other languages.  We find that such directions can indeed be consistently assigned in a way that yields valid multi-headed dependency parses. The resulting parses in English appear reasonably linguistically plausible, though differing in style from CoNLL-style parses of the same sentences; we discuss the differences.  
\end{abstract}

@@

As discussed by \newcite{kuhlmann-nivre-2006,gomezrodriguez-nivre-2013}, one might impose various requirements on the parse graph:
\begin{itemize}[noitemsep]
\item {\sc Single-Head}: each word has $\leq 1$ parent
\item {\sc Acyclic}: there are no directed cycles
\item {\sc Connected}: each pair of words has a undirected path between them
\item {\sc Reachable}: each word can be reached from 0 by a directed path ($\Rightarrow$ {\sc Connected})
\item {\sc Planar}: edges may not ``cross''\footnote{That is, if there are edges between $i,j$ and between $u,v$, where $i < u < j$, then {\sc Planar} requires $i \leq v \leq j$.}
\end{itemize}
It is common to impose all of these requirements at once, leading to a {\em projective dependency parser} that produces projective trees rooted at 0.\footnote{{\sc Projective} basically means {\sc Planar} + {\sc Reachable}.  Note: It does not prevent 0 from having multiple children.}  However, parsing algorithms can be devised that relax any or all of the requirements \cite{gomezrodriguez-nivre-2013}.

@@

Graph representations of syntactic and semantic structure have been widely considered of late \cite{demarneffe-manning-2008,ivanova-et-al-2012,AMR-2013,oepen-et-al-2014}.

A few past NLP papers have explored multi-headed dependency parsing \cite{buchkromann-2006,mcdonald-pereira-2006,sagae-tsujii-2008,gomezrodriguez-nivre-2013}.  Unfortunately, there seem to be no annotated corpora in this form other than the Danish Dependency Treebank \cite{kromann-2003}; these papers  sometimes converted corpora from other formats such as HPSG.

@@

It seems at first that no one has worked out annotation conventions for {\em projective} multi-headed dependency parsing.  However, this is only half-true.  Link Grammar \cite{SleatorTemperly91} is a grammar-based formalism for projective dependency parsing with {\em undirected} links.  It produces undirected connected planar graphs.  Annotation conventions are implicit in the detailed lexicon for the Link Grammar Parser,\noteJE{check caps}%
\footnote{\url{http://www.abisource.com/projects/link-grammar/dict/introduction.html}.  The 122 link types in the English lexicon are documented at \url{http://www.abisource.com/projects/link-grammar/dict/summarize-links.html}.} 
which specifies for every word a constraint on the {\em sequence} of labeled leftward and rightward edges attached to it.  As remarked by \newcite{eisner-2000-iwptbook}, this is analogous to dependency grammar's use of head automata to constrain a word's sequence of left and right children.  For example, in {\em ``The boy that Jill skipped with fell down,''} the word {\em with} uses a lexical entry that requires it to link to a governing verb to the left, an extracted object farther to the left, and nothing to the right.  Each entry has a hand-assigned cost in \{0,1,2\} and the parser finds the parse of minimum total cost.
% Information about cost is in this message from Linas Vepstas on 2014-02-01: 
%   https://groups.google.com/d/msg/link-grammar/eeJw1Ofgc9U/diqPYSwuFfoJ

@@

% REMOVED AT REVIEWER SUGGESTION
% However, working directly with the link grammar lexicon format is somewhat tricky. We also feared
However, we feared that there would not be a feasible solution---because of errors in the lexicon or linguistically unnatural word sequences not anticipated by the grammar designers.  In this case it would be unclear how to relax our constraints.
